---
title: 'Penkrot et al., 2018 Geosphere U1419 general cluster analysis '
author: "Michelle Penkrot"
date: "3/19/2018"
output:
  word_document: default
  html_document: default
---

##Code Information
This code will import physical property data and normalized scanning XRF elemental concentrations (NMS normalized; Lyle et al., 2012) from Integrated Ocean Drilling Site U1419, and then perform both mixture-model clustering and heirarchical clustering on these data. See here for more details about this drilling location: http://iodp.tamu.edu/scienceops/expeditions/alaska_tectonics_climate.html. 

This file will  produce 2-3 principle components and one to five clusters, which is the general model output. The user should be aware that running the full model (this code) takes considerable computing time. It is optimized to use parallel processing on four cores. Running the mclust routine for five clusters with two and three principle components can take up to 24 hours. 

Note that the cluster number (e.g., Cluster 1, Cluster 2, etc.) produced by he code is arbitrary, so we renamed the clusters post-modeling in rank order so that Cluster 1 indicates the most common cluster and Cluster 5 the least common.

The mclust portion of the following code is modified from R scripts that were originally written and kindly supplied by Karl Ellefsen (2015, personal communication). See Ellefsen, K. J., Smith, D. B., & Horton, J. D. (2014). A modified procedure for mixture-model clustering of regional geochemical data. Applied Geochemistry, 51, 315–326. https://doi.org/10.1016/j.apgeochem.2014.10.011 
```{r setup, include=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

##Load Packages
```{r load_packages, message=FALSE, results='hide'}
library(knitr)
library(rgr)
library(ggplot2)
library(psych)
library(parallel)
library(rrcov)
library(mclust)
library(lsr)
library(parallel)
```

##Create functions
```{r functions, message=FALSE}
CalcSampleClusters <- function( theData, nPDFs, sampleSize, sampleSpace, nIter )
{
  require( mclust, quiet=TRUE )
  
  theLogLikelihoods <- vector( mode="numeric" )
  nErrors <- 0
  for( i in 1:nIter ) {
    S <- sample( sampleSpace, size = sampleSize )
    
    clusterResult <- tryCatch( Mclust( theData, nPDFs, modelNames=c("VVV"), initialization=list(subset=S) ), error=function(e){ e } )
    
    if( inherits( clusterResult, "error" ) ) {
      nErrors <- nErrors + 1
    } else {
      theLogLikelihoods <- append( theLogLikelihoods, clusterResult$loglik )
      if( max( theLogLikelihoods ) == clusterResult$loglik ) {
        bestClusterResult <- clusterResult
      }
      if( min( theLogLikelihoods ) == clusterResult$loglik ) {
        worstClusterResult <- clusterResult
      }
    }
  }
  
  return( list( theLogLikelihoods=theLogLikelihoods, bestClusterResult=bestClusterResult,
                worstClusterResult=worstClusterResult, nErrors=nErrors ) )
}
```

##Load Data
Loads the data from a csv file. Separates data based on type (i.e. physical property or scanning XRF elemental). The physical property data are centered and scaled (z-score), and an isometric log ratio transformation (ILR) is performed on the elemental data to open them. 

NOTE: See comments below to set the code to run on the full mud-diamict dataset. It is set here to run only on the diamict-only portion of the data.
```{r load data,message=FALSE}
data<-read.csv("../raw_data/2018-03-20_U1419-Penkrot_Geosphere-2018-data.csv")
data<-data[509:6490,] #cuts the data down to the diamict-only portion of the core; comment this line if you want to run the analyses on the full mud-diamict data set.
#lith<-data$Mud_Diamict_code
lith<-data$Diamict_only_code #comment this line and uncomment previous line ifyou want to run the analyses on the full mud-diamict data set.
depth<-data$CCFS_A

physprops<-data[c("b_star","NGR","MS")]
elements<-data[c("Al","Ca","Rb","Zr","K","Si")]

physprops_scaled<-scale(physprops, center=TRUE,scale=TRUE) # Centers and scales the physical property data
elements_ilr<-ilr(as.matrix(elements)) # Performs isometric log ratio transformation to open XRF data
```

##Select Data for Models
Select which data to include in cluster analysis. This study ran the model 3 times, with physical property and scanning XRF elemental data inputs (run name: G1), only physical property inputs (run name: G2) and only scanning XRF data (run name G3). The optimal data set was G3 and is used in the paper.
```{r select elements}
#run.name<-"G1" # Change for each cluster model run
#run.name<-"G2" # Change for each cluster model run
run.name<-"G3" # Change for each cluster model run

#data_transformed<-as.matrix(cbind(physprops_scaled,elements_ilr)) #phys props & scanning XRF data; G1
#data_transformed<-as.matrix(cbind(physprops_scaled)) #phys props only; G2
data_transformed<-as.matrix(cbind(elements_ilr)) #scanning XRF only; G3
```

##Data Processing
Performs a robust principle component analysis on the input data to remove outliers. These principle component results are used in the cluster analysis rather than the raw data. The alpha value lets you select which percentage of the data to be excluded as outliers. Follows this method: Martin Maechler, Peter Rousseeuw, Christophe Croux, Valentin Todorov, Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel Koller, Eduardo L. T. Conceicao and Maria Anna di
  Palma (2016). robustbase: Basic Robust Statistics R package version 0.92-7. URL http://CRAN.R-project.org/package=robustbase
```{r robustPCA}
# select alpha(s): 0.98, 0.96, and 0.92, which will exclude 2%, 4%, and 8% of the data as outliers, respectively
alpha <- c( 0.98)
outputDirectories <- c( "../produced_data/") # Change to appropriate output directory

  mcdResult <- covMcd( data_transformed, alpha=alpha )
  robustIlrCenter <- mcdResult$center
  robustIlrCov <- mcdResult$cov
  centeredIlrCoefs <- data_transformed -  rep.int( 1, nrow( data_transformed ) ) %o% robustIlrCenter
  svdResult <- svd( robustIlrCov )
  robustEigenvectors <- svdResult$u
  robustEigenvalues <- svdResult$d   # namely, robust variances
  robustPCs <- centeredIlrCoefs %*% robustEigenvectors
  plot(robustEigenvalues) # Scree plot
  save(robustIlrCenter, robustEigenvectors, robustEigenvalues, robustPCs,
       file=paste( outputDirectories,run.name,"_PrinComp.dat", sep="" ) )
``` 

## Model-based clustering (MClust)
Performs model clustering on robust principle component results. The number of principles components (nPCs) can be changed in line 121 and the number of clusters (nPDFs) can be changed in line 123. This study ran the model cluster analysis for both 2 & 3 PCs and 2-5 clusters. 
```{r Model Clustering, eval=F, echo=T}

nWorkers <- 4                   # should be <= 4 on my machine, i.e., number of processor cores
cl <- makeCluster( nWorkers )
clusterSetRNGStream( cl, 123 )

outputDirectories <- c("../produced_data/") #Change to appropriate output directory
                       
                       for( j in 1:length(outputDirectories) ) {
                       #for( j in 1:1 ) {
                       
                       load( paste( outputDirectories[j],run.name,"_PrinComp.dat", sep="" ) )
                       
                       nRows <- nrow( robustPCs )
                       sampleSize <- as.integer( 0.75 * nRows )
                       sampleSpace <- 1:nRows
                       
                       for( nPCs in 3:3) { #uses 2 & 3 principle components
                       
                       for( nPDFs in 5:5 ) { #creates 2-5 clusters
                       
                       clusterOutputDirectory <- paste( outputDirectories[j], run.name,"_modelclust_",nPCs, "-PCs__", nPDFs, "-PDFs/", sep="" )
                       dir.create(clusterOutputDirectory)
                       
                       nIter <- 100 + ( nPDFs - 2 ) * 150 + ( nPCs - 4 ) * 30
                       nIterPerWorker <- as.integer( nIter / nWorkers )
                       cat( sprintf( "No. of principal components:  %3d    No. of pdfs: %3d    No. of iter: %3d    No. of iter per worker: %3d\n",
                       nPCs, nPDFs, nIter, nIterPerWorker ) )
                       
                       tmpResult <-clusterCall( cl, CalcSampleClusters, robustPCs[,1:nPCs], nPDFs, sampleSize, sampleSpace, nIterPerWorker  )
                       
                       theLogLikelihoods <- tmpResult[[1]]$theLogLikelihoods
                       bestClusterResult <- tmpResult[[1]]$bestClusterResult
                       worstClusterResult <- tmpResult[[1]]$worstClusterResult
                       nErrors <- tmpResult[[1]]$nErrors
                       for( i in 2:nWorkers ) {
                       
                       theLogLikelihoods <- append( theLogLikelihoods, tmpResult[[i]]$theLogLikelihoods )
                       
                       if( bestClusterResult$loglik < tmpResult[[i]]$bestClusterResult$loglik )
                       bestClusterResult <- tmpResult[[i]]$bestClusterResult
                       
                       if( worstClusterResult$loglik > tmpResult[[i]]$worstClusterResult$loglik )
                       worstClusterResult <- tmpResult[[i]]$worstClusterResult
                       
                       nErrors <- nErrors + tmpResult[[i]]$nErrors
                       }
                       
                       save( theLogLikelihoods, bestClusterResult, worstClusterResult, nErrors,
                       file=paste( clusterOutputDirectory,"data.dat", sep="" ) )
                       
                       cat( sprintf( "No. of errors: %3d\n", nErrors ) )
                       }
                       }
                       }
                       
                       stopCluster( cl )
```

###Combines model clustering results into one matrix.
```{r model clustering results}

pc<-c(2,2,2,2,3,3,3,3)
pdf<-c(2,3,4,5,2,3,4,5)

mclust_results<-as.data.frame(matrix(0,nrow(data_transformed),length(pc)))
for(i in 1:length(pc)){
load(paste("../produced_data/",run.name,"_modelclust_",pc[1],"-PCs__",pdf[1],"-PDFs/data.dat",sep="")) 
bestclusterresult<-bestClusterResult$classification
mclust_results[,i]<-bestclusterresult
colnames(mclust_results)[i]<-paste("mclust_",pc[i],"PCs_",pdf[i],"Clusters",sep="")
}

load(paste("../produced_data/G3_modelclust_3-PCs__5-PDFs/data.dat",sep="")) 
bestclusterresult<-bestClusterResult$classification
mclust_results<-bestclusterresult
```

##Hierarchical-based clustering (hclust)
Performs Hierarchical cluster analysis using the hclust function from the R stats package. Hierarchical clustering produces a dendrogram based on similariities of the downcore data. The dendrogram is cut off at heights that produces 2-5 clusters. 
```{r Hierarchical Clustering}
distance2<-dist(as.data.frame(robustPCs[,1:2]),method="euclidean") # uses 2 principle components
MyTree2<-hclust(distance2, method='ward.D2')
Cut2PC_2CL<-as.data.frame(cutree(MyTree2,k=2)) # cuts hclust tree at 2 clusters
Cut2PC_3CL<-as.data.frame(cutree(MyTree2, k=3))
Cut2PC_4CL<-as.data.frame(cutree(MyTree2, k=4))
Cut2PC_5CL<-as.data.frame(cutree(MyTree2, k=5)) # cuts hclust tree at 5 clusters
                       
distance3<-dist(as.data.frame(robustPCs[,1:3]),method="euclidean") # uses 3 principle components
MyTree3<-hclust(distance3, method='ward.D2')
Cut3PC_2CL<-as.data.frame(cutree(MyTree3,k=2)) # cuts hclust tree at 2 clusters
Cut3PC_3CL<-as.data.frame(cutree(MyTree3, k=3))
Cut3PC_4CL<-as.data.frame(cutree(MyTree3, k=4))
Cut3PC_5CL<-as.data.frame(cutree(MyTree3, k=5)) # cuts hclust tree at 5 clusters
                       
hclust_results<-cbind(Cut2PC_2CL,Cut2PC_3CL,Cut2PC_4CL,Cut2PC_5CL,Cut3PC_2CL,Cut3PC_3CL,Cut3PC_4CL,Cut3PC_5CL)
colnames(hclust_results)<-c("hclust_2PCs-2Clusters","hclust_2PCs-3Clusters","hclust_2PCs-4Clusters","hclust_2PCs-5Clusters","hclust_3PCs-2Clusters","hclust_3PCs-3Clusters","hclust_3PCs-4Clusters","hclust_3PCs-5Clusters")

hclust_results<-(Cut3PC_5CL)
colnames(hclust_results)<-c("hclust_3PCs-5Clusters")
```

##Statistical Analysis of Results
Statistical parameters used to validate cluster results through comparison with downcore changes in observed lithofacies. Chi-squared test, Cramer's V-value, F-measure and Rand Index are calculated. This study only discusses the Chi-squared test and Cramer's V-value results. 
```{r Statistical Validation}
results<-cbind(depth,mclust_results,hclust_results)
r<-length(results)-1  
c<-4
statistics<-matrix(0,r,c) # r=number of models, c=number of statistical tests
for (i in 1:r){
  q<-results[,i+1]
  tbl<-table(lith,q)
  chi<-chisq.test(tbl)
  cv<-cramersV(tbl)
  x<-chi$statistic
  y<-chi$parameter
  z<-chi$p.value
  statistics[i,]<-cbind(x,y,z,cv)
}
colnames(statistics)<-c("X-square","Df","p-value","Cramers V Value")
nCols <- ncol(results)
rownames(statistics)<-colnames(results[,2:nCols])
```

##Save Model Output
Saves clustering results from both model and hierarchical clustering for 2-3 PCs and 2-5 clusters, and statistical parameter results as .csv files.
```{r Save all results}
write.csv(results,file=paste("../produced_data/",toString(run.name),"_clusteringresults.csv",sep="")) # Saves Hierarchical and model clustering results
write.csv(statistics,file=paste("../produced_data/",toString(run.name),"_statistics.csv",sep="")) # Saves statistical validation results
```

##Create plots of cluster models
Plots of downcore distributions of cluster results for both model and hierarchical clustering, 2-3 PCs and 2-5 clusters. Plots go in "plot" RProject folder. 
```{r Plot cluster results}
titles<-colnames(results)
resultsd<-results[!duplicated(results$depth),]
nColsL <- ncol(resultsd)
for(i in 2:nColsL){
  cluster<-factor(resultsd[,i])
  p<-ggplot(resultsd, aes(x=resultsd$depth,y=resultsd[,i],fill=cluster))+geom_bar(stat="identity",size=3,width=0.1) +
    scale_fill_manual(values=rainbow(n=length(unique(resultsd[,i])))) + coord_flip() + scale_x_reverse() + 
    labs(x="CCSF-A, meters",y=toString(colnames(resultsd)[i])) 
  print(p)
  ggsave(filename=paste("../plot/",toString(run.name),"_",toString(colnames(resultsd)[i]),".pdf",sep=''),height=12,width=3)
  dev.off()
}
```

